---
title: "vectorizedCorrelations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vectorizedCorrelations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(vectorizedCorrelations)
```

`vectorizedCorrelations` computes correlations using only basic arithmetic operations and (in some cases) matrix transposition native to `R`. The function `matrix_cor()` can compute two types of correlation coefficients: [*pearson*](https://en.wikipedia.org/wiki/Pearson_correlation_coefficient) correlation, and [*kendall*](https://en.wikipedia.org/wiki/Kendall_rank_correlation_coefficient) rank correlation.

## Pearson correlation

You can get the pearson correlation coefficient between the columns of two matrices. 

```{r}
set.seed(1120)
a = matrix(rnorm(5*3, mean = 10, sd = 1), nrow=5, ncol=3)
b = 2*a + rnorm(5*3, mean = 1, sd = 0.1)
matrix_cor(a, b, "pearson")
```

In the output above, the entry [*i*, *j*] corresponds to the correlation between column *i* in matrix *a* and column *j* in matrix *b*. So, for example, the entry [2, 3] is the correlation between the second column of *a* and the third column of *b*.

Compare this result to the one from `stats::cor()`

```{r}
stats::cor(a, b, method = "pearson")
```

You can also use two vectors:

```{r}
set.seed(1120)
a = rnorm(5*3, mean = 10, sd = 1)
b = 2*a + rnorm(5*3, mean = 5, sd = 2)
matrix_cor(a, b, "pearson")
```

or one vector and one matrix

```{r}
set.seed(1120)
a = matrix(rnorm(10*4, mean = 10, sd = 1), nrow=10, ncol=4)
b = 2*a[, 1]
matrix_cor(a, b, "pearson")
# Or
matrix_cor(b, a, "pearson")
```

It is also possible to get the correlation matrix for a single matrix by passing only the first matrix and the method:

```{r}
set.seed(1120)
a = matrix(rnorm(10*4, mean = 10, sd = 1), nrow=10, ncol=4)
matrix_cor(a, method="pearson")
```

In most cases, the output in `matrix_cor()` has the same format as that of `stats::cor()`. The differences occur when you pass a vector in matrix format:

```{r}
set.seed(20)
a = matrix(rnorm(7), nrow=7, ncol=1)
b = matrix(rnorm(7), nrow=7, ncol=1)
matrix_cor(a, b, method = "pearson")
cor(a, b, method = "pearson")
```

Here, `matrix_cor()` outputs a vector of length 1, while `stats::cor()` outputs a matrix of dimensions (1, 1).

## Kendall rank correlation

To get the kendall rank correlation, you can use `method = "kendall"`. 

```{r}
set.seed(854)
a = matrix(rnorm(10*10), nrow=10, ncol=10)
b = matrix(rnorm(10*10), nrow=10, ncol=10)
matrix_cor(a, b, "kendall")
```

Note that in this case `matrix_cor()` is more limited: it only outputs a vector where the element *i* corresponds to the kendall correlation between column *i* of the first matrix and the column *i* of the second matrix.

Even so, it is still possible to pass two vectors:

```{r}
set.seed(1120)
a = rnorm(5*3, mean = 10, sd = 1)
b = 2*a + rnorm(5*3, mean = 5, sd = 2)
matrix_cor(a, b, "kendall")
```

or one vector and one matrix:

```{r}
set.seed(1120)
a = matrix(rnorm(10*4, mean = 10, sd = 1), nrow=10, ncol=4)
b = 2*a[, 1]
matrix_cor(a, b, "kendall")
matrix_cor(b, a, "kendall")
```







